---
title: "Class09: Candy Mini Project"
author: "Raneem Kassar (PID:A17803411)"
format: pdf
toc: true
---

## Background
Today we are taking sa wee detour to analyze a fun data-set with the most useful analysis method we have learned thus far.

## 2 Importing candy data

```{r}
candy <- read.csv("https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv", row.names = 1)
head(candy)
```

## 2.1 What is in the dataset?

> Q1. How many differeent candy types are in this dataset?

```{r}
nrow(candy)
```
There are 85 different candy types.

> Q2. How many fruity candy types are in this dataset?

```{r}
sum(candy$fruity)
```
There are 38 fruity candy types are in this dataset.

## 2.2 What is your favorite candy?

> Q3. What is your favorite candy (other than Twix) in the dataset and what is it’s winpercent value?

```{r}
candy["Reese's Peanut Butter cup",]$winpercent
```
My favorite candy in the dataset is Reese's Peanut Butter Cup and it's win percent value is 84.18029.

> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat",]$winpercent
```
 The winpercent value for "Kit Kat" is 76.7686.

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars",]$winpercent
```
 The winpercent value for "Tootsie Roll Snack Bars is 49.6535.
 
```{r}
skimr::skim(candy)
```

 > Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Yes, the variable that looks most different in scale is winpercent. Most of the other columns are on a scale that range as 0 to 1. For example, chocolate and caramel only use 0 and 1 values, also, sugarpercent and pricepercent are percentile values as 0 pr 1. However, winpercent is measured as a percentage from 0 to 100%.
 
 > Q7. What do you think a zero and one represent for the candy$chocolate column?

For the candy$chocolate column, a zero would represent the candy contains no chocolate, while 1 represents the candy contains chocolate.

## 3 Exploratory analysis

> Q8. Plot a histogram of winpercent values using both base R and ggplot2.

```{r}
hist(candy$winpercent)
```

I plotted a histogram of winpercent values using both base R and ggplot2.

```{r}
library(ggplot2)

ggplot(candy) +
  aes(winpercent) +
  geom_histogram(bins=20, fill="steelblue")
```

> Q9. Is the distribution of winpercent values symmetrical?

The distribution of winpercent values seems to be not perfecltly symmetrical and more leaning right skewed.

> Q10. Is the center of the distribution above or below 50%?

```{r}
mean(candy$winpercent)
```

```{r}
summary(candy$winpercent)
```

The center of distribution is around 50%. The mean is slightly above 50% at 50.31676, but the median is slightly below at 47.83. Overall, the distribution is centered close to 50%.

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
choc.ind <- as.logical(candy$chocolate)
choc.candy <- candy[choc.ind, ]
choc.win <- choc.candy$winpercent
mean(choc.win)
```

```{r}
fruit.ind <- as.logical(candy$fruity)
fruit.candy <- candy[fruit.ind,]
fruit.win <- fruit.candy$winpercent
mean(fruit.win)
```

On average, chocolate is higher ranked than fruity candy as chocolate's mean is 60.92153, and fruit candy is 44.11974.

> Q12. Is this difference statistically significant?

```{r}
t.test(choc.win, fruit.win)
```

This difference is statistically significant because the welch two sample t-test gives a p-value of 2.871e-08, which is much smaller than 0.05. There is strong evidence that chocolate candies and fruity candies have different average winpercent values.

## 4 Overall Candy Rankings

> Q13. What are the five least liked candy types in this set?

```{r}
library(dplyr)

candy |>
  arrange(winpercent) |>
  select(winpercent) |>
    head(5)
```

The five least liked candy types in this set are Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbusters.

```{r}
x <- c(5,10,1)
sort(x)
```

```{r}
head(candy [ order(candy$winpercent), ], 5)
```

> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
library(dplyr)

candy |>
  arrange(desc(winpercent)) |>
  select(winpercent) |>
  head(5)
```

The five all time favorite candy types in this data set are Reese's Peanut Butter cup, Reese's Miniatures, Twix, Kit Kat, and Snickers.

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r, fig.height=15, fig.width=10}
ggplot(candy) +
  aes(winpercent, rownames(candy)) +
  geom_col()
```
This barplot shows the overall candy rankings based on winpercent values. Each candy is on the y axis while the win percentage is on the x axis.

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r, fig.height=15, fig.width=10}
ggplot(candy) +
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col()
```
I used the reorder() function to sort the candy names by their winpercent values.

## 4.0.1 Time to add some useful color


```{r, fig.height=15, fig.width=10}
mycols <- rep("black", nrow(candy))

## Chocolate candy in chocolate color
mycols [ as.logical(candy$chocolate) ] <- "chocolate"
# candy bars in brown
mycols [as.logical(candy$bar) ] <- "brown"
# fruity candy in pink
mycols [as.logical(candy$fruity) ] <- "pink"


ggplot(candy) + 
 aes(winpercent, 
     reorder(rownames(candy), winpercent)) +
  geom_col(fill = mycols) 
```


> Q17. What is the worst ranked chocolate candy?

The worst ranked chocolate candy is Sixlets.

> Q18. What is the best ranked fruity candy?

The best ranked fruit candy according to the plot is Starburst.

## 5 Taking a look at pricepercent

```{r}
ggplot(candy) +
  aes(winpercent, 
      pricepercent,
        label=rownames(candy)) +
  geom_point(col=mycols) +
  geom_text()
```

We can fixthe label overplotting with an add-on package called **ggrepel** and it's `geom_text_repel()`


```{r}
# fruity candy in pink
mycols [as.logical(candy$fruity) ] <- "red" #"pink"
```

```{r}
library(ggrepel)

ggplot(candy) +
  aes(winpercent, 
      pricepercent,
        label=rownames(candy)) +
  geom_point(col=mycols) +
  geom_text_repel(col = mycols, size = 3.3, max.overlaps = 5)
```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

The candy that is highest ranked in terms of winpercent is Reese's minatures with a winpercent of 81.86626 with a low pricepercent value.

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

The top 5 most expensive candy types in the dataset are Nik L Nip, Nestle Samrtes, Ring Pop, Hershey;s Krackel, and Hershey's Milk Chocolate. The least popular amogn these is Nik L Nip with a winpercent value at 22.44534.

## 6 Exploring the correlation structure

We can calculate the pair-wise correlation of all our columns 

```{r}
cij <- cor(candy)

library(corrplot)
corrplot(cij)
```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

The two variables that most anti-correlated in this data set are chocolate and fruity.

> Q23. Use your corrplot result to make a prediction: which variables do you expect will have the largest contributions (i.e. loadings) to PC1 (i.e., drive the most separation between candies along the first principal component)?

The variables I expect will have the largest contributions to PC1 are chocolate and fruity. This due to to the variables strongly seperate candy types due to chocolate and fruit candies being negatively correlated. Also, variables like bar and winpercent will contribute to PC1 due to chocolate tends tend to be more popular and expensive.

## 7 Principal Component Analysis

In this case we want to be set `scale=TRUE` argument for `prcomp()`because we have on variable, `win percent` that is on a very different scale than all others and would otherwise dominate our PCA results.

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

First major result figure is the "score plot" of PC1 vs PC2 - how the different candy are related to each other on our new PC axis: 

```{r}
ggplot(pca$x) +
  aes(PC1, PC2, label=row.names(pca$x)) +
  geom_point(col=mycols) +
  geom_text_repel(size=3.3, col=mycols, max.overlaps = 7)
```

The second major results figure from PCA is the so-called "loadings plot" 

```{r}
ggplot(pca$rotation)+
  aes(PC1, 
      reorder( row.names(pca$rotation), PC1))+
  geom_col()
```

> Q24. Complete the code to generate the loadings plot above. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you? Where did you see this relationship highlighted previously?

The original variables picked up strongly by PC1 in the positive direction are chocolate, bar, winpercent, and pricepercent. These make sense to me due to chocolate and candy bars tend to be more popular and expensive in the dataset. I saw this relationship highlighted previously in the corrplot.

## 8 Summary

> Q25. Based on your exploratory analysis, correlation findings, and PCA results, what combination of characteristics appears to make a “winning” candy? How do these different analyses (visualization, correlation, PCA) support or complement each other in reaching this conclusion?

Based on my exploratory analysis, correlation results, and PCA, a "winning" candy appears to be one that is chocolate-based, often a bar and may include peanuts, caramel, or wafer. These candies tend to have higher winpercent values. Also, the ranked bar plots showed the top candies such as Reese's, Kit Kat, Twix, and Snickers, which are all chocolate candy. Different analyses support or complement each other in reaching this conclusion through variables such as chocolate, pricepercent, and winpercent being positively related. Finally, PCA supported the same pattern by showing that PC1 separated chocolate bar from fruity candies.


