---
title: "Class 5: Data Viz with ggplot2"
author: "Raneem (PID: A17803411)"
format: pdf
toc: true
---

## Background

There are many graphics systems in R for making plots and figures. These include so-called *'base R" graphics* like the 'plot()' function and on packages like **ggplot2**.

Let's compare how we make a simple figure with these two systems: 

We can ue the in-built 'cars' dataset:

```{r}
head( cars )
```

```{r}
plot(cars)
```

Before I can use ggplot2 i need to install it on my computer.
To do this we can use the function 'install.packages("ggplot2")'

> **N.B.** We never run 'install.packages()' in our quarto doc (we run it once only in our R console) as it would re-install the package every time we render our quarto report.

Once installed we need to load up the package into our R brain: 

```{r}
library(ggplot2)
```

The main function in the **ggplot2** package is called 'ggplot()'

```{r}
ggplot(cars)
```
Every ggplot has at least 3 layers: 


- the **data** (a data.frame of the stuff we want to plot)
- the **aes**thetics (how the data maps to the plot).
- the ***geom** layer (how you want the plot drawn, e.g. points, lines, etc.)

```{r}
ggplot(cars) +
  aes(x=speed, y=dist ) +
  geom_point()
```

## Add some custom features

Let's add a trend line that shows the relationship between speed and distance.










```{r}
ggplot(cars) +
  aes(x=speed, y=dist ) +
  geom_point() +
  geom_smooth(method = "lm", se=FALSE) +
  theme_bw() + 
  labs(title="Stoping Dist of old cars",
       x="Speed (MPH)",
       y ="Distance(ft)")
```


> Q. Can you argue make the 'geom_smooth()' function produce a linear straight line fit to the data and turn off the "gray" error region. 

---

## Gene expression figure

Import the data to plot

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

```{r}
sum(genes$State == "up")
```
A useful new function in this context is the 'table()' function:

```{r}
table(genes$State)
```

My first plot attempt

```{r}
ggplot(genes) + 
  aes(Condition1, Condition2, col=State) + 
  geom_point() +
  scale_color_manual(values=c("purple", "gray", "orange"))+
  theme_bw() +
  labs(x="No drug", y= "Drug",
       title = "Expression changes upon GLP-1 inhibitor treatment")
```

## Going further

Here we read the famous gapminder dataset:

```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
head(gapminder)
```

> Q. How many entries (i.e. rows) are in this dataset?

```{r}
nrow(gapminder)
```
> Q. How many different"country" entries are in this dataset?


```{r}
length( table(gapminder$country) )
```
```{r}
length( unique(gapminder$country) )
```

Let's make our first plot of the entire dataset:

Plot of "gdpPercap" vs "lifeExp" colored by "continent"

```{r}
p <- ggplot(gapminder) +
  aes(gdpPercap, lifeExp, col=continent) +
  geom_point(alpha=0.3)
```

```{r}
p
```

I can add more layers to 'p'
```{r}
p + 
  facet_wrap(~continent)
```

Make a plot for year 1977 and 2007 only (not all the years in the dataset).

> Q. First use the **dplyr** package and the 'filter()' function from that package to extract the rows from the year 2007.

```{r, message=FALSE}
library(dplyr)
```

```{r}
g07 <- filter(gapminder, year == 2007)
g07 <- filter(gapminder, year == 1977)
g <- filter(gapminder, year == 2007 | year ==1977)

```

```{r}
ggplot(g) +
aes(gdpPercap, lifeExp, col=continent, size=pop) +
  geom_point() +
  facet_wrap(~year)
```

> Q. Make a histogram of lifeExp colored by continent (try use 'fill=continent' or 'col=continent')

```{r}
ggplot(gapminder) +
  aes(x = lifeExp, fill = continent) +
  geom_histogram()
```


> Q. Make a histogram of lifeExp faceted by continent

```{r}
ggplot(gapminder) +
  aes(lifeExp, fill = continent) +
  geom_histogram() + 
facet_wrap(~ continent)
```





