---
title: "Class 12: Pt2 (Population Analysis)"
author: "Raneem Kassar (PID: A17803411)"
format: pdf
---

> Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

```{r}
data <- read.table("https://bioboot.github.io/bggn213_W19/class-material/rs8067378_ENSG00000172057.6.txt",
                   header = TRUE)
table(data$geno)
aggregate(exp ~ geno, data = data, median)
```
The rs8067378 ORMDL3 expressing file sample sizes were 108 A/A, 233 A/G, and 121 G/G.  The median expression values were 31.24847	for A/A, 25.06486	for A/G, and 20.07363 for G/G.

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

```{r}
boxplot(exp ~ geno,
        data = data,
        xlab = "Genotype",
        ylab = "ORMDL3 Expression",
        main = "ORMDL3 Expression by rs8067378 Genotype")
```

Yes, rs8067378 appears to affect ORMDL3 expression because individuals with the G/G genotype has lower ORMDL3 expression than invidiuals with A/A genotype.