---
title: "Class7: Machine Learning 1"
author: "Raneem Kassar (PID: A17803411)"
format: pdf
toc: true
---

## Background

Today we will explore some corre machine learning methods that are very popular in bioinformatics. These include **clustering** and **dimensionallity reduction**.


## K-means clustering 

The main function in "base" R for K-means clustering is called `kmeans()`

Before we go too deep let's make up some "simple" data that we can cluster and know if we are getting a good answer or not. To do this we can use `Rnorm()` function:

```{r}
hist( rnorm(10000, mean =3) )
```

```{r}
x<- c(rnorm(30, -3), rnorm(30, +3) )
z <- cbind(x=x, y=rev(x))
plot(z)
```




Now we can run `kmeans()` on this input `z` and see what the results look like.

```{r}
km <- kmeans(z, centers = 2)
km
```

```{r}
attributes(km)
```

> Q. How many points are in each cluster?

```{r}
km$size
```


> Q. What "component of your result object details cluser assignment/membership?

```{r}
km$cluster
```

> Q. What "component of your result object details cluster center?

```{r}
km$centers
```


>Q. plot `z` colored by the kmeans cluster assignment and add cluster centers as blue points

```{r}
plot(z, col = c("red", "blue") )

```

```{r}
plot(z, col= km$cluster)
points(km$centers, col="blue", pch=15)
```

> Q. Run a K-means clustering and plot the results asking for 4 clusters (K=4)?

```{r}
km4 <- kmeans(z, centers = 4)
plot(z, col=km4$cluster)
points(km4$centers, col="black", pch=15)
```

> **N.B.** You need to tell K-means the number of clusters (ie set `centers=2`)!!

One approach is to try different values for `centers` and then pick the best...

```{r}
ans <- NULL
for(i in 1:10) {
km <- kmeans(z, centers=i)
ans <- c(ans,km$tot.withinss)
}

plot(ans, type="o", 
     xlab="Number of Clusters",
     ylab="Total Sum of Squares Distance")
```

# Hierarchical Clustering

The main function in "base" R for Hierarchical Clustering is called `hclust()`

This function does not take your "raw" data for clustering. You must first build a "distance matrix" from your data and pass this as input to `hclust()`


```{r}
d <- dist(z)
hc <- hclust(d)
hc
```

There isa bespoke `plot()` methdo for `hclust()` result objects.

```{r}
plot(hc)
abline(h=8, col="red")
```

Once we have our `hclust` object (our "tree" of "cluster dendrogram) we can *"cut"* the tree to reveal the clustering pattern.

```{r}
cutree(hc, h=8)
```


```{r}
cutree(hc, k=4)
```

> Q. Make a plot of `z` with your hclust results (i.e. colored by cluster membership)

```{r}
grps <- cutree(hc, k=2)
plot(z, col=grps)
```


## Principal Component Analysis (PCA)

PCA is a dimensionallity reduction method that is popular for revealing patterns in complex datasets

## Analysis of UK food data

Let's look at some data on the eating habits of folks from the UK to see if there are patterns and trends that have some regionss being distinct from others.

## Data Import

The data is made avaliable in CSV format so we can use `read.csv()` function for import R

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
x
```

```{r}
rownames(x) <- x[,1]
x <- x[,-1]
```


> Q1. How many rows and columns are in your new data frame named x? What R functions could you use to answer this questions?

```{r}
dim(x)
```

There are 17 rows and 4 columns.

## Tidy the data

Fix anything that went wrong with data import.

>Q2. Which approach to solving the ‘row-names problem’ mentioned above do you prefer and why? Is one approach more robust than another under certain circumstances?

```{r}
head(x)
dim(x)
```

I prefer using read.csv(url, row.names =1) as it solves the row names problem during the data import and the approach is less likely to cause issues due to the first column being treated as row names. However, manually setting the rownames(x) <- x[,1] and then removing the x can be more robust than the other if you want to inspect the dat first and confirm the column names.

## Exporatory Analysis

Make some plots to help make sense of obvious trends...

## PCA

```{r}
barplot(as.matrix(x), beside=T, col=rainbow(nrow(x)))
```

```{r}
barplot(as.matrix(x), beside= FALSE, col=rainbow(nrow(x)))
```


> Q3. Changing what optional argument in the above barplot() function results in the following plot?

Changing beside= T to beside= FALSE. 



```{r}
library(tidyr)

x_long <- x |> 
  tibble::rownames_to_column("Food") |> 
  pivot_longer(cols = -Food, 
               names_to = "Country", 
               values_to = "Consumption")

dim(x_long)
head(x_long)
```

```{r}
library(ggplot2)
```


```{r}
ggplot(x_long) +
  aes(x = Country, y = Consumption, fill = Food) +
  geom_col(position = "stack") +
  theme_bw()
```

>. Q4: Changing what optional argument in the above ggplot() code results in a stacked barplot figure.

Changing the optional argument position int the geom_col() function from "dodge" to "stack" results in a stacked barplot.

>. **Q5:** We can use the pairs() function to generate all pairwise plots for our countries. Can you make sense of the following code and resulting figure? What does it mean if a given point lies on the diagonal for a given plot?

```{r}
pairs(x, col = rainbow(nrow(x)), pch = 16)
```

The pairs() function makes scatter plots comparing every pair of countries, each point represents a food type. If a point lies on the diagonal line for a given plot this means that the food has similar consumption value in both countries being compared. Points far from the diagonal show foods consumed differently between countries.


```{r}
library(pheatmap)
pheatmap(as.matrix(x))
```

> Q6. Based on the pairs and heatmap figures, which countries cluster together and what does this suggest about their food consumption patterns? Can you easily tell what the main differences between N. Ireland and the other countries of the UK in terms of this data-set?

Based on the pairs plot and heat map, Wales, England, and Scotland cluster more closely together while Ireland is more separate This suggests that they have more similar food consumption patterns while Ireland is different. It is a bit easier to view Ireland differing from other countries but still difficult to identify main differences.

> **Key Point**: Even relatively small dataets can prove challenging to interpret.

## PCA to the rescue

The main function in "base" R for PCA is called `prcomp()`. This function wants the "observations to be rows and the "variables" to be columns.

So here we need to take the transpose of our `x` input object

```{r}
pca <- prcomp(t(x))
summary(pca)
```



```{r}
# Create a data frame for plotting
df <- as.data.frame(pca$x)
df$Country <- rownames(df)

# Plot PC1 vs PC2 with ggplot
ggplot(pca$x) +
  aes(x = PC1, y = PC2, label = rownames(pca$x)) +
  geom_point(size = 3) +
  geom_text(vjust = -0.5) +
  xlim(-270, 500) +
  xlab("PC1") +
  ylab("PC2") +
  theme_bw()
```

> Q7. Complete the code below to generate a plot of PC1 vs PC2. The second line adds text labels over the data points.

The blanks are PC1 and PC2, which creates a plot comparing the two components.




The returned 'pca' object has components that we can use to make oru main result figures:

```{r}
attributes(pca)
```
The main result figure from this analysis is called a **"PC score plot" or "ordenation plot"** "PC plot" or "PC1 vs PC2 plot".

This plot shows gow samples (jn this case countries) relate to each other alogn our new PC axis.

This is our new "reduced-dimensional space". In this case 2 dimesnions, PC1 and PC2, that capture most of the variance in the original 17 dimensional data-set.

```{r}
library(ggplot2)

ggplot(pca$x) + 
  aes(PC1, PC2) + 
  geom_point()
```

```{r}
mycols <- c("orange", "red", "blue", "darkgreen")

ggplot(pca$x) + 
  aes(PC1, PC2) + 
  geom_point(col=mycols)
```


> Q8. Customize your plot so that the colors of the country names match the colors in our UK and Ireland map and table at start of this document.

Customization are shown below.

```{r}
pca$x

ggplot(pca$x) + 
  aes(PC1, PC2, label=row.names(pca$x)) + 
  geom_point(col=mycols) +
  geom_text(size=3, vjust=2, col=mycols)
```

```{r}
ggplot(pca$rotation) +
  aes(PC1, reorder(row.names(pca$rotation), PC1) ) +
  geom_col()
```

> Q9: Generate a similar ‘loadings plot’ for PC2. What two food groups feature prominantely and what does PC2 mainly tell us about?

```{r}
ggplot(pca$rotation) +
  aes(PC2, reorder(row.names(pca$rotation), PC2) ) +
  geom_col()
```
The two food spots that feature prominently in PC2 are soft_drinks on the positive side and fresh_potatoes which is on the negative side. This tells us that the contrast between countries associated more with soft drink consumption versus fresh potato consumption.












